---
title: "Small area population estimates, England, Wales & Scotland - 2011 on"
author: "Nick Bailey, UBDC"
date: "18/01/2022"
output: 
  html_document:
    code-folding: hide
---

## LSOA & DZ populations (2011 geographies) - 2011 on

Reads data from UK/devolved government spreadsheets containing small area population estimates for Lower Super Output Areas (LSOAs) in England and Wales and Datazones (DZs) in Scotland, and combines them into a single file. Units are the 2011-based LSOA/DZ geographies. Extracts total population and three banded age groups (0-15; 16-64; and 65+). Several of the source .csv files have data for individual years of age, but others are banded into 5-year age groups. Only the .csv file for 2011 LSOAs has banding which does not align with the three age groups selected. In this case, an estimate is made of the number of 15-year olds (assumed as one fifth of the number 15-19) so the same age groups can be constructed. 

Data downloaded (18 Jan 2022) from: 

* https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datasets/lowersuperoutputareamidyearpopulationestimates and 
* https://www.nrscotland.gov.uk/statistics-and-data/statistics/statistics-by-theme/population/population-estimates/small-area-population-estimates-2011-data-zone-based/time-series

Resulting .csv file contains estimated mid-year population (popln; popln_0_15; popln_16_64; popln_65_up) for 41,729 LSOAs/DZs (lsoa_code) for England, Wales and Scotland for each year (year) from 2011 onwards. 

```{r setup}
# knitr::opts_chunk$set(echo = FALSE)

# load packages
pacman::p_load(feather, here, tidyverse, readxl)

```


## England & Wales

34,753 LSOAs in England & Wales. 

```{r data ew}

## 2020 - with age bands
df_pop_ew_20 <- read_excel(here("sape_ew",
                                  "sape23dt2mid2020lsoasyoaestimatesunformatted.xlsx"),
                             sheet = "Mid-2020 Persons",
                             skip = 4) %>%
  mutate(popln_0_15 = rowSums(select(., `0`:`15`)),
         popln_16_64 = rowSums(select(., `16`:`64`)),
         popln_65_up = rowSums(select(., `65`:`90+`))) %>%
  select(lsoa_code = `LSOA Code`,
         popln = `All Ages`,
         popln_0_15,
         popln_16_64,
         popln_65_up) %>%
  filter(grepl("E01", lsoa_code)|
           grepl("W01", lsoa_code)) %>% # remove rows with national total, etc.
  mutate(year = 2020)

## 2019 updated May 2023
df_pop_ew_19 <- read_excel(here("sape_ew", 
                                "SAPE22DT2-mid-2019-lsoa-syoa-estimates-unformatted.xlsx"),
                             sheet = "Mid-2019 Persons", 
                             skip = 4) %>% 
  mutate(popln_0_15 = rowSums(select(., `0`:`15`)),
         popln_16_64 = rowSums(select(., `16`:`64`)),
         popln_65_up = rowSums(select(., `65`:`90+`))) %>%
  select(lsoa_code = `LSOA Code`,
         popln = `All Ages`,
         popln_0_15,
         popln_16_64,
         popln_65_up) %>%
  filter(grepl("E01", lsoa_code)|
           grepl("W01", lsoa_code)) %>% 
  mutate(year = 2019)

## 2018
df_pop_ew_18 <- read_excel(here("sape_ew", 
                                "SAPE21DT1a-mid-2018-on-2019-LA-lsoa-syoa-estimates-formatted.xlsx"),
                             sheet = "Mid-2018 Persons", 
                             skip = 4) %>% 
  mutate(popln_0_15 = rowSums(select(., `0`:`15`)),
         popln_16_64 = rowSums(select(., `16`:`64`)),
         popln_65_up = rowSums(select(., `65`:`90+`))) %>%
  select(lsoa_code = `Area Codes`,
         popln = `All Ages`,
         popln_0_15,
         popln_16_64,
         popln_65_up) %>%
  filter(grepl("E01", lsoa_code)|
           grepl("W01", lsoa_code)) %>% 
  mutate(year = 2018)

## 2017
df_pop_ew_17 <- read_excel(here("sape_ew", 
                                  "SAPE20DT1-mid-2017-lsoa-syoa-estimates-formatted.xls"),
                             sheet = "Mid-2017 Persons", 
                             skip = 4) %>% 
  mutate(popln_0_15 = rowSums(select(., `0`:`15`)),
         popln_16_64 = rowSums(select(., `16`:`64`)),
         popln_65_up = rowSums(select(., `65`:`90+`))) %>%
  select(lsoa_code = `Area Codes`,
         popln = `All Ages`,
         popln_0_15,
         popln_16_64,
         popln_65_up) %>%
  filter(grepl("E01", lsoa_code)|
           grepl("W01", lsoa_code)) %>% 
  mutate(year = 2017)

## 2016
df_pop_ew_16 <- read_excel(here("sape_ew", 
                                  "SAPE20DT1-mid-2016-lsoa-syoa-estimates-formatted.xls"),
                             sheet = "Mid-2016 Persons", 
                             skip = 4) %>% 
  mutate(popln_0_15 = rowSums(select(., `0`:`15`)),
         popln_16_64 = rowSums(select(., `16`:`64`)),
         popln_65_up = rowSums(select(., `65`:`90+`))) %>%
  select(lsoa_code = `Area Codes`,
         popln = `All Ages`,
         popln_0_15,
         popln_16_64,
         popln_65_up) %>%
  filter(grepl("E01", lsoa_code)|
           grepl("W01", lsoa_code)) %>% 
  mutate(year = 2016)

## 2015
df_pop_ew_15 <- read_excel(here("sape_ew", 
                                  "SAPE20DT1-mid-2015-lsoa-syoa-estimates-formatted.xls"),
                             sheet = "Mid-2015 Persons", 
                             skip = 4) %>% 
  mutate(popln_0_15 = rowSums(select(., `0`:`15`)),
         popln_16_64 = rowSums(select(., `16`:`64`)),
         popln_65_up = rowSums(select(., `65`:`90+`))) %>%
  select(lsoa_code = `Area Codes`,
         popln = `All Ages`,
         popln_0_15,
         popln_16_64,
         popln_65_up) %>%
  filter(grepl("E01", lsoa_code)|
           grepl("W01", lsoa_code)) %>% 
  mutate(year = 2015)

## 2014
df_pop_ew_14 <- read_excel(here("sape_ew", 
                                  "SAPE20DT1-mid-2014-lsoa-syoa-estimates-formatted.xls"),
                             sheet = "Mid-2014 Persons", 
                             skip = 4) %>% 
  mutate(popln_0_15 = rowSums(select(., `0`:`15`)),
         popln_16_64 = rowSums(select(., `16`:`64`)),
         popln_65_up = rowSums(select(., `65`:`90+`))) %>%
  select(lsoa_code = `Area Codes`,
         popln = `All Ages`,
         popln_0_15,
         popln_16_64,
         popln_65_up) %>%
  filter(grepl("E01", lsoa_code)|
           grepl("W01", lsoa_code)) %>% 
  mutate(year = 2014)

## 2013
df_pop_ew_13 <- read_excel(here("sape_ew", 
                                  "SAPE20DT1-mid-2013-lsoa-syoa-estimates-formatted.xls"),
                             sheet = "Mid-2013 Persons", 
                             skip = 4) %>% 
  mutate(popln_0_15 = rowSums(select(., `0`:`15`)),
         popln_16_64 = rowSums(select(., `16`:`64`)),
         popln_65_up = rowSums(select(., `65`:`90+`))) %>%
  select(lsoa_code = `Area Codes`,
         popln = `All Ages`,
         popln_0_15,
         popln_16_64,
         popln_65_up) %>%
  filter(grepl("E01", lsoa_code)|
           grepl("W01", lsoa_code)) %>% 
  mutate(year = 2013)

## 2012
df_pop_ew_12 <- read_excel(here("sape_ew", 
                                  "SAPE20DT1-mid-2012-lsoa-syoa-estimates-formatted.xls"),
                             sheet = "Mid-2012 Persons", 
                             skip = 4) %>% 
  mutate(popln_0_15 = rowSums(select(., `0`:`15`)),
         popln_16_64 = rowSums(select(., `16`:`64`)),
         popln_65_up = rowSums(select(., `65`:`90+`))) %>%
  select(lsoa_code = `Area Codes`,
         popln = `All Ages`,
         popln_0_15,
         popln_16_64,
         popln_65_up) %>%
  filter(grepl("E01", lsoa_code)|
           grepl("W01", lsoa_code)) %>% 
  mutate(year = 2012)

## 2011
#  NB estimates made for 0-15 and 16-64 due to banding in source data
df_pop_ew_11 <- read_excel(here("sape_ew", 
                                  "mid-2011-lsoa-quinary-estimates.xls"),
                             sheet = "Mid-2011 Persons", 
                             skip = 3) %>%   # note diff to later years
  mutate(popln_0_15 = rowSums(select(., `0-4`:`10-14`)) + (`15-19`/5),
         popln_16_64 = rowSums(select(., `15-19`:`60-64`)) - (`15-19`/5),
         popln_65_up = rowSums(select(., `65-69`:`90+`))) %>%
  select(lsoa_code = `Area Codes`,
         popln = `All Ages`,
         popln_0_15,
         popln_16_64,
         popln_65_up) %>%
  filter(grepl("E01", lsoa_code)|
           grepl("W01", lsoa_code)) %>% 
  mutate(year = 2011)

# combine
df_pop_ew <- df_pop_ew_20 %>% 
  rbind(df_pop_ew_19) %>% 
  rbind(df_pop_ew_18) %>% 
  rbind(df_pop_ew_17) %>% 
  rbind(df_pop_ew_16) %>% 
  rbind(df_pop_ew_15) %>% 
  rbind(df_pop_ew_14) %>% 
  rbind(df_pop_ew_13) %>% 
  rbind(df_pop_ew_12) %>% 
  rbind(df_pop_ew_11)

```


## Scotland
6976 DZs in Scotland. Note that the Scottish data go back to 2001 on the same set of (2011-based) DZ boundaries. 
May 2023: a new, consistently formatted set of .xlsx files with single year of age now available making automating possible. 


```{r data sc}

# make list of years for file names
ls_files_sc <- unlist(seq(2001, 2021, 1))

# read .xlsx files and reformat
df_pop_sc <- ls_files_sc %>%
  map_dfr(~ read_excel(here("sape_sc", paste0("sape-", .x, ".xlsx")),
                       sheet = "Persons",
                       skip = 3) %>% mutate(year = .x)) %>% 
  mutate(popln_0_15 = rowSums(select(., `Age 0`:`Age 15`)),
         popln_16_64 = rowSums(select(., `Age 16`:`Age 64`)),
         popln_65_up = rowSums(select(., `Age 65`:`Age 90 and over`))) %>%
  select(lsoa_code = `Data zone code`,
         popln = `Total population`,
         popln_0_15,
         popln_16_64,
         popln_65_up, 
         year) %>%
  filter(grepl("S01", lsoa_code))

```




## Combine files

Combine files and tidy. 

```{r data all}

# combine ew and sc
df_pop <- df_pop_ew %>% 
  rbind(df_pop_sc) 

```


Check that, for each LSOA, total population is the sum of the different age bands. Result should be zero. 

```{r check 1}

df_pop %>% 
  mutate(check = popln - rowSums(select(., popln_0_15:popln_65_up))) %>% 
  summarise(check_sum = sum(check**2)) # use square so negatives can't cancel positives

```

```{r data all tidy}

# combine ew and sc
df_pop <- df_pop %>% 
  pivot_longer(contains("popln"), names_to = "popln", values_to = "count")

```

Check the number of LSOAs is the same each year. Result should be 41,729 from 2011 onwards; earlier years just Scotland (and current version, 2021 also).

```{r check 2}

df_pop %>% 
  filter(popln == "popln") %>% 
  group_by(year) %>% 
  summarise(n_lsoa = n())

```

## Save result

```{r save}

# write 2011 years on for GB
write_csv(df_pop[df_pop$year >= 2011, ], 
          here("data", "lsoa_dz_popln_2011_on.csv"))

# write 2011 years on for GB
write_feather(df_pop[df_pop$year >= 2011, ], 
              here("data", "lsoa_dz_popln_2011_on.feather"))

# Scotland only - all years
temp <- df_pop %>% 
  filter(grepl("S01", lsoa_code)) 

write_csv(temp, here("data", "dz_popln_2001_on.csv"))

```

